From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-21 Pi+/paper/finite.tex
Finish Coxeter section

%% Coxeter presentation of $\Sn$ $\eqv$ $\Lehmer[n]$ $\eqv$ $\Aut[\Fin[n]]$

@@

This construction is called a Coxeter presentation. Writing it formally, we get
the following definition for $\Sn$. We define a binary relation $\cox$ on
$\List[\Fin[n]]$.

@@

@@

To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. The $\cox*$ relation as defined, does not produce a
well-behaved rewriting system due to the symmetric closure. Throwing out reflexivity and symmetry, we observe that the
right hand sides of the relations are strictly smaller than the left hand sides, in terms of the lexicographical
ordering on words in $\Fin[n]$. By directing the relation from left to right, we would be able to prove a termination
property out of the box.

A well-behaved rewriting system should also be confluent, however, it is not at all obvious from this definition. Due to
overlapping reduction rules, there are many possible critical pairs, such as

@@

For confluence, all critical pairs should converge. The pairs above can be resolved, but there are many examples where
they do not. The following is a simple example.

@@

Because of this counterexample, the relations have to be changed. We propose a long form of the relations as a new
presentation, where we allow $\cancel$ and $\swap$ reductions anywhere inside the list, called $\longcancel$ and
$\longswap$, but replace $\braid$ with a more general $\longbraid$ relation.

The directed $\longcox$ relation is defined formally with the following generators.

@@

Despite the increased complexity of the relations, the rewriting system $(\List[\Fin[n]],\longcox*)$ can be shown to
satisfy confluence, that is, the Church-Rosser (diamond) property. Further, we can show that this system is strongly
normalising, by producing a unique normal form that doesn't reduce any further. We follow the terminology
of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\longcox$ is (locally) confluent. For every span $\coxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
  \end{enumerate}
\end{proposition}

Another property we're interested in is well-foundedness of the relation $\longcox*$.
\todo{lexicographic order, accessibility predicate and well-foundedness}

In the theory of group presentations, the Coxeter relations are generally presented in their short form~\todo{citation},
but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}. The long form of the
Coxeter relations are unwieldy and difficult to prove properties about by induction, and have little connection to the
broader scope of computational group theory. However, we can prove that the closures of the two forms of the relations
are equivalent, and work up to this equivalence!

@@

By strong normalisation, we get a unique choice function that produces a normal form for terms of $\List[\Fin[n]]$.
Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have the same normal form.

20. Now, to prove the equivalence of 9, on one side we have Aut(Fin n), and
on the other, equivalence classes of List (Fin n) wrt to the usual Coxeter
group, we would define two functions from and back. The easiest way to
define a function out of the group presentation is to define them on the
equivalence classes' representatives. But to do that, we have to know how
do these representatives look like exactly.

The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.

We will now prove the group structure of $\Sn$.

@@

\begin{proposition}
  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
\end{proposition}

\subsection{Lehmer Codes}

21. Enter: Lehmer codes.
TODO diagram with List (Fin n) being divided into equivalence classes, and
Lehmer codes being an image of immersion, being the normal form
(representative) in each class.

\subsection{Permutations}~\label{subsec:permutations}

\vc{This is obvious, maybe add something more here.}

\subsection{Lehmer codes}~\label{subsec:lehmer}

>>>>

@@

This construction is called a Coxeter presentation. Writing it formally, we define a binary relation $\cox$ on
$\List[\Fin[n]]$.

@@

The type $\Sn$ is then defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.

@@

To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. A well-behaved rewriting system has to be strongly
normalising and confluent. Throwing out reflexivity and symmetry, we observe that the right hand sides of the relations
are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$. Thus, by
directing the relation from left to right, we would be able to prove a termination property out of the box. For a system
to be confluent, all critical pairs, that is, terms with overlapping possible reduction rules, have to converge. For
example, the pairs below converge.

@@

Unfortunately, in the system defined, this is not true for all critical pairs. One example is the following.

@@

\subsection{Rewriting via Coxeter}

Because of the counterexample discussed, the relations have to be changed. In this section, we will formally define a
rewriting system partially based on the Coxeter relation, prove that it has good properties (confluence and strong
normalisation).

The reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$. We
fix the problem discussed previously by changing $\braid$ to be a slightly more general relation. However, we do not
want to lose the connection to the theory of group presentations, where Coxeter relations are a standard
notion~\todo{citation}. To make the result broadly applicable, even though we use the reduction system for computation,
we prove that its reflexive-transitive closure is indeed equivalent to the conguence closure of the standard Coxeter
relation $\cox$.

\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}

Formally, the directed $\longcox$ relation is defined with the following generators.

@@

Generators $\longcancel$ and $\longswap$ correspond directly to the generators of $\cox*$ and can be visualised in the
same way as before. The generator $\longbraid$ uses a helper $n \downf k$, which produces the sequence
\(k + n, k + n - 1 \ldots, n + 1\).

\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
  \begin{align*}
    n \downf 0       & \defeq \nil                       \\
    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
  \end{align*}
\end{definition}

As a sequence of transpositions, it moves the element $k - 1$ places. Expressed in terms of the braiding diagram, for
$n = \dots$ and $k = \dots$, it has the following form:

\[
  \begin{tikzpicture}
    \pic[local bounding box=my braid,braid/.cd,
      number of strands = 6,
      thick]
    {braid={s_5, s_4, s_3, s_2, s_1}};
  \end{tikzpicture}
\]

Now, the remaining constructor $\longbraid$ moves a transposition around such sequence, as in the diagram below.

Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 2$.

Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox*)$ has the properties
we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising,
because it produces a unique irreducible normal form. Existence of the normal form is implied by the well-foundedness of
the relation -- informally, it does not contain any infinite decreasing sequences. We follow the terminology
of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results formally.~\vc{check!}

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item $\longcox$ is (locally) confluent. For every span $\coxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
    \item $\longcox*$ is well-founded. With $<$ the lexicographic ordering on $\List[\Fin[n]]$, and $w$ being
          $<$-accessible if every $v < w$ is $<$-accessible, we have that every $w$ is $<$-accesible.
    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
  \end{enumerate}
\end{proposition}

The long form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and have
little connection to the broader scope of computational group theory. However, we can prove that the closures of the two
forms of the relations are equivalent, and work up to this equivalence!

@@

By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have
the same normal form. We state a few important properties that $\normf$ satisfies.

\begin{proposition}
  \leavevmode
  \begin{enumerate}
    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
  \end{enumerate}
\end{proposition}

\vc{Remark that $w \cox* v$ is not a proposition.}

Finally, we prove the group structure of $\Sn$, and show that it is indeed a group presentation, given by a quotient of
the free group $F(\Fin[n])$.

@@

\begin{proposition}
  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
\end{proposition}

\subsection{Lehmer Codes}~\label{subsec:lehmer}

To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
representatives. We know that these are the unique normal forms in the set-quotient given by $q \comp \normf$, but now
we will describe what these representatives look like exactly, using Lehmer codes.

\todo{diagram with List (Fin n) being divided into equivalence classes, and Lehmer codes being an image of immersion,
  being the normal form (representative) in each class.}
